---
title: "Analysis code for prediction of BK polyomavirus-associated nephropathy"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

This code reproduces the model development, internal validation, external validation, performance measures, and figures reported in the manuscript.

Analyses begin from two analysis-ready datasets:

- `df_au`: Australian derivation cohort
- `df_nz`: New Zealand external validation cohort

Individual-level ANZDATA Registry data cannot be distributed publicly because access is subject to Registry governance and ethics requirements. Consequently, source-data extraction and dataset-construction code are not included in this repository.


# Load packages
```{r}
library(tidyverse)
library(pROC)
library(patchwork)
library(jtools)
```


# Performance derivation
## Apparent AUC
```{r}



model_au <- glm(formula = event ~ ageattransplant + BMI + CNI + dgf + donorsourcecode + egfr + era + ethnicity + gendercode + induction_tcelldep + antimetabolite + prednisolone + rejection + tercile + time_period, family = binomial(link = "cloglog"),
    data = df_au)

predictions_au <- predict(model_au, type = "link")

# Area under the curve
auc_apparent <- roc(df_au$event, predictions_au)$auc
auc_apparent

ci.auc(auc_apparent)

```

## Apparent Brier
```{r}

p <- 1 - exp(-exp(predictions_au))
brier <- mean((df_au$event - p)^2)
brier
```


## ROC curve
```{r}
roc_au <- roc(
  response = df_au$event,
  predictor = predictions_au
)

roc_df_au <- data.frame(
  specificity = roc_au$specificities,
  sensitivity = roc_au$sensitivities
)

ci_vals <- ci.auc(roc_au)

ci_text <- paste0(
  round(ci_vals[1], 2),
  "–",
  round(ci_vals[3], 2)
)

p_deriv <- ggplot(roc_df_au, aes(x = 1 - specificity, y = sensitivity)) +
  geom_line(size = 1, colour = "#BC3C29FF") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "#34495e"
  ) +
  coord_equal() +
  labs(
    title = "Derivation cohort (Australia)",
    x = "1 - Specificity",
    y = "Sensitivity"
  ) +
  annotate(
     "label",
    x = 0.65,
    y = 0.20,
    label = paste0(
      "AUC = ", round(auc(roc_au), 2), " (95% CI: ", ci_text, ")"
    ),
    size = 5,
    fill = "white",
    colour = "#34495e",
    linewidth = 0.5
  ) +
  theme_nice()

p_deriv

```


## Optimism adjusted AUC
```{r}

set.seed(2026)

B <- 500
n <- nrow(df_au)

bootstrap_results <-
  tibble(
    b = 1:B,
    idx = map(b, ~ sample(seq_len(n), replace = TRUE))
  ) %>%
  mutate(

    # Bootstrap dataset
    data_boot = map(idx, ~ df_au[.x, ]),

    # Fit model in bootstrap sample
    fit_boot = map(data_boot, ~ glm(
      model_au,
      family = binomial(link = "cloglog"),
      data = .x
    )),

    # Apparent AUC in bootstrap sample
    auc_boot_app = map2_dbl(
      fit_boot, data_boot,
      ~ roc(.y$event, predict(.x, type = "link"))$auc
    ),

    # Test AUC in original AU data
    auc_boot_test = map_dbl(
      fit_boot,
      ~ roc(df_au$event, predict(.x, newdata = df_au, type = "link"))$auc
    ),

    optimism_auc = auc_boot_app - auc_boot_test,


    # ----- Calibration slope -----

    # Apparent slope in bootstrap sample
    slope_boot_app = map2_dbl(
      fit_boot, data_boot,
      ~ {
        lp_boot <- predict(.x, type = "link")
        cal_boot <- glm(
          event ~ lp_boot,
          family = binomial(link = "cloglog"),
          data = .y
        )
        coef(cal_boot)[2]
      }
    ),

    # Test slope in original AU data
    slope_boot_test = map_dbl(
      fit_boot,
      ~ {
        lp_test <- predict(.x, newdata = df_au, type = "link")
        cal_test <- glm(
          event ~ lp_test,
          family = binomial(link = "cloglog"),
          data = df_au
        )
        coef(cal_test)[2]
      }
    ),

    optimism_slope = slope_boot_app - slope_boot_test
  )

bootstrap_results <- bootstrap_results %>%
  select(b, auc_boot_app, auc_boot_test, optimism_auc, slope_boot_app, slope_boot_test, optimism_slope)


mean_optimism_auc <- mean(bootstrap_results$optimism_auc)
auc_corrected <- auc_apparent - mean_optimism_auc

auc_corrected
```

## Optimism adjusted calibration 
```{r}
cal_app_au <- glm(
  event ~ predictions_au,
  family = binomial(link = "cloglog"),
  data = df_au
)

intercept_app <- coef(cal_app_au)[1]
slope_app <- coef(cal_app_au)[2]

mean_optimism_slope <- mean(bootstrap_results$optimism_slope)

slope_corrected <- slope_app - mean_optimism_slope

slope_app
slope_corrected

```

## Apparent calibration curve
```{r}

p_deriv_calib <- df_au %>%
  mutate(pred_prob = 1 - exp(-exp(predictions_au))) %>% 
    mutate(decile = ntile(pred_prob, 10)) %>%
  group_by(decile) %>%
  summarise(
       predicted = mean(pred_prob),
    observed = mean(event),
    events = sum(event),
    n = n(),
    .groups = "drop") %>%  
  ggplot(aes(x = predicted, y = observed)) +
  geom_point(size = 3, col = "#34495e") +
  geom_smooth(se = F, span = 1, col = "#BC3C29FF", linewidth = 0.75) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", col = "#34495e") +
  labs(
    x = "Mean Predicted Risk",
    y = "Observed Event Rate",
    title = "Derivation cohort (Australia)"
  ) +
  coord_equal() +
  theme_nice() +
  coord_equal(
    xlim = c(0, 0.03),
    ylim = c(0, 0.03)
  )

p_deriv_calib
```


# Performance validation
## AUC
```{r}

predictions_nz <- predict(model_au, newdata = df_nz, type = "link")
roc_nz <- roc(df_nz$event, predictions_nz)

auc_nz <- auc(roc_nz)

ci_nz <- ci.auc(roc_nz, method = "delong")


auc_nz
ci_nz

```


## Brier 
```{r}
p_nz <- predict(model_au, newdata = df_nz, type = "response")
brier_nz <- mean((df_nz$event - p_nz)^2)
brier_nz

```



## ROC curve
```{r}
roc_df_nz <- coords(
  roc_nz,
  x = "all",
  ret = c("specificity", "sensitivity"),
  transpose = FALSE
) %>% 
  as.data.frame() %>% 
  mutate(
    fpr = 1 - specificity,
    tpr = sensitivity
  )

p_valid <- ggplot(roc_df_nz, aes(x = fpr, y = tpr)) +
  geom_line(size = 1, colour = "#BC3C29FF") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "#34495e"
  ) +
  labs(
    x = "1 - Specificity",
    y = "Sensitivity",
    title = "External validation cohort (New Zealand)"
  ) +
  annotate(
     "label",
    x = 0.65,
    y = 0.20,
    label = paste0(
      "AUC = ", round(auc_nz, 2),
      " (95% CI: ", round(ci_nz[1], 2),
      "–", round(ci_nz[3], 2), ")"
    ),
    size = 5,
    fill = "white",
    colour = "#34495e",
    linewidth = 0.5
  ) +
  theme_nice()

p_valid

p_deriv + p_valid
```


## Calibration
```{r}
cal_model <- glm(
  event ~ predictions_nz,
  family = binomial(link = "cloglog"),
  data = df_nz
)

summary(cal_model)

df_nz_cal <- df_nz %>%
  mutate(
    lp = predictions_nz,
    pred = 1 - exp(-exp(lp))
  )

df_cal_plot <- df_nz_cal %>%
  mutate(
    decile = ntile(pred, 10)
  ) %>%
  group_by(decile) %>%
  summarise(
    mean_pred = mean(pred),
    obs = mean(event),
    n = n(),
    .groups = "drop"
  )

p_valid_calib <- ggplot(df_cal_plot, aes(x = mean_pred, y = obs)) +
  geom_point(size = 3, col = "#34495e") +
  geom_smooth(se = F, span = 1, col = "#BC3C29FF", linewidth = 0.75) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", col = "#34495e") +
  labs(
    x = "Mean Predicted Risk",
    y = "Observed Event Rate",
    title = "Validation cohort (New Zealand)"
  ) +
  coord_equal() +
  theme_nice()  +
  coord_equal(
    xlim = c(0, 0.03),
    ylim = c(0, 0.03)
  )



p_valid_calib

p_deriv_calib + p_valid_calib

```
